Temporal ordering and registration of cross-sectional imaging data

### 2014-05-19 image_analysis_paper/pnas_draft2.tex
movie code, edited mutant code, etc.

Here, we show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data. 
%
Based on manifold learning algorithms for high-dimensional data (such as images), these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition. 
%
Here we adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We demonstrate these techniques using three data sets from a study of {\it Dropophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
%
The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate our methods. 
%
In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
The third data set, shown in Figure~\ref{subfig:raw_data3}, consists of data collected from both wild type and mutant embryos.
%
We will demonstrate that these algorithms can not only extract stereotypic dynamics, but also separate data sampled from distinct developmental paths.

@@

Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using our techniques. 
%
For example, the images contain size variability. 
%
In this example, the size variation is relatively small relative to the developmental changes. 
%
However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or scaling symmetries.

@@

%
During this developmental time span, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
%
dpERK (red) again appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
During this time window, germband elongation causes cells from the ventral side of the embryo to move towards the poles of the embryo, and then wrap around to the dorsal side.
%
At the end of this process, cells which were originally on the ventral side have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side.

@@

Vector diffusion maps can not only temporally order data, but can also uncover other sources of variability.
%
Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by time.
%
However, this is not always the case.
%
As an example, Figure~\ref{subfig:raw_data3} shows a data set consisting of $42$ embryos undergoing cellularization during the third hour of development. 
%
Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
%
In this data set, $21$ of the embryos are wild type, and $21$ of the embryos are an sna null mutant.
%
We therefore expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.

@@

We can compute the average trajectory over each of the two curves to gain insight into the different developmental dynamics.
%
The average over the blue curve, shown in Figure~\ref{subfig:average_wt_data3}, shows the emergence of two distinct dpERK peaks, as expected.
%
The average over the red curve, shown in Figure~\ref{subfig:average_mut_data3}, reveals one more concentrated dpERK peak.
%
This is consistent with previously known dynamics.
%
At early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, whereas at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and sna null mutants develop only a single peak.

\begin{figure*}[t]
\raisebox{8.8cm}{{\figtextfont A}}
\includegraphics[height=9cm]{raw_data3}
\begin{minipage}[b]{8.5cm}
\raisebox{5.3cm}{{\figtextfont B}}
\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}
\raisebox{1.3cm}{{\figtextfont C}}
\includegraphics[width=8.4cm]{wt_trajectory}
\raisebox{1.3cm}{{\figtextfont D}}
\includegraphics[width=8.4cm]{mut_trajectory}
\end{minipage}
\caption{Mutant and wild type {\xfigtextfontit Drosophila}. {\xfigtextfontit (A)} Images of cellularizing {\xfigtextfontit Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half of the images are of sna null mutant embryos. {\xfigtextfontit(B)} Data from {\xfigtextfontit A}, projected onto the first two vector diffusion maps coordinates. The two distinct trajectories are highlighted in blue and red. {\xfigtextfontit (C)} Average trajectory of the dpERK signal over the blue curve in {\xfigtextfontit B}. {\xfigtextfontit (D)} Average trajectory of the dpERK signal over the red curve in {\xfigtextfontit B}.}
\customlabel{fig:data3}{5}
\customlabel{subfig:raw_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
\end{figure*}

@@

\bibliographystyle{pnas}
\bibliography{background_reading/references,../../references/references}

>>>>

Here, we show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data. 
%
Based on manifold learning algorithms for high-dimensional data (such as images), these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition. 
%
Here we adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We demonstrate these techniques using three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
%
The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate our methods. 
%
In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
The third data set, shown in Figure~\ref{subfig:raw_data3}, consists of data collected from both wild type and mutant embryos.
%
We will demonstrate that these algorithms can not only extract stereotypic dynamics, but also separate data sampled from distinct developmental paths.

@@

Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using our techniques. 
%
For example, the images contain size variability. 
%
In this example, the size variation is relatively small relative to the developmental changes. 
%
However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or scaling symmetries, or consider additional components.

@@

%
During this developmental time span, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
%
dpERK (red) again appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
During this time window, germband elongation causes cells from the ventral side of the embryo to move towards the poles of the embryo, and then wrap around to the dorsal side.
%
At the end of this process, cells which were originally on the ventral side have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side (shown in the late images in Figure~\ref{subfig:averaged_data2}).

@@

Vector diffusion maps can not only temporally order data, but can also uncover other sources of variability.
%
Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by time.
%
However, this is not always the case.
%
As an example, Figure~\ref{subfig:raw_data3} shows a data set consisting of $42$ embryos undergoing cellularization during the third hour of development. 
%
Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
%
In this data set, $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side, instead of in two bimodal peaks.
%
We therefore expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.

@@

We can compute the average trajectory over each of the two curves to gain insight into the different developmental dynamics.
%
The average over the blue curve, shown in Figure~\ref{subfig:average_wt_data3}, shows the emergence of two distinct dpERK peaks, as expected.
%
The average over the red curve, shown in Figure~\ref{subfig:average_mut_data3}, reveals one more concentrated dpERK peak.
%
This is consistent with previously known dynamics.
%
At early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, whereas at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and the mutant embryos develop only a single peak.
%
We have therefore demonstrated that vector diffusion maps can not only recover dynamics, but also distinguish different expression patterns and developmental trajectories.

\begin{figure}[t]
%\raisebox{8.8cm}{{\figtextfont A}}
%\includegraphics[height=9cm]{raw_data3}
%\begin{minipage}[b]{8.5cm}
\raisebox{6cm}{{\figtextfont B}}
\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}
\raisebox{2cm}{{\figtextfont C}}
\includegraphics[width=8.4cm]{wt_trajectory}
\raisebox{2cm}{{\figtextfont D}}
\includegraphics[width=8.4cm]{mut_trajectory}
%\end{minipage}
\caption{Mutant and wild type {\xfigtextfontit Drosophila}. {\xfigtextfontit (A)} Images of cellularizing {\xfigtextfontit Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half of the images are of sna null mutant embryos. {\xfigtextfontit(B)} Data from {\xfigtextfontit A}, projected onto the first two vector diffusion maps coordinates. The two distinct trajectories are highlighted in blue and red. {\xfigtextfontit (C)} Average trajectory of the dpERK signal over the blue curve in {\xfigtextfontit B}. {\xfigtextfontit (D)} Average trajectory of the dpERK signal over the red curve in {\xfigtextfontit B}.}
\customlabel{fig:data3}{5}
\customlabel{subfig:raw_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
\end{figure}

@@

%\bibliographystyle{pnas}
%\bibliography{background_reading/references,../../references/references}
